---
title: "Colon cancer analysis before removal of arrays with poor quality"
output:
  html_document:
    df_print: paged
---

```{r, echo = FALSE, results='hide', message =FALSE}
memory.limit(9999999999)

library(kableExtra)
library(knitr)
library(purrr)
library(dplyr)

load("B:/OneDrive/Documents/Microarray data/02_Normalized data/01_Colon/DataInfo.RData")

```

# Data summary

Table below summarizes GEO projects used in the analysis:


```{r, echo = FALSE, resutls = "asis", message =FALSE}


kable(bind_rows(DataInfo)[,-6], 
      caption = "Information about included projects") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(1, bold = T) # 
```

To further characterize the data use we look at the number of samples belonging to each project, using specific platform and representing a different phenotype:

```{r, echo = FALSE, resutls = "asis", message =FALSE}
source("ShapeData.R")

kable(PhenoSummary(PhenoData), 
      caption = "Number of samples per project, platform and phenotpye") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(dim(PhenoSummary(PhenoData))[1], bold = T) %>% # format last row
  column_spec(1, bold = T) #

```

# Data preprocessing

## Quality analysis

CEL. files belonging to relevant projects were read in using ReadAffy function and visually inspected for  array quality was assessed before further processing.

```{r}
#source("QualityAnalysis.R")
#load("B:/OneDrive/Documents/Microarray data/02_Normalized data/01_Colon/RawData.RData")

#purrr::map2(RawExp, DataInfo, QualityAnalysis)
```



```{r, echo = FALSE, resutls = "hide", message =FALSE}
library(lemon)

load("B:/OneDrive/Documents/Microarray data/02_Normalized data/01_Colon/NormData.RData")
load("B:/OneDrive/Documents/Microarray data/02_Normalized data/01_Colon/GeneExpression_median.RData")

# Box plots
source("Visualize.R")
par(mfrow=c(1,3))
BP_RMA_BG = BoxPlot(RMA_BG,
                    Logaritmic = "y",
                    Title="RMA background correction")
BP_RMA =BoxPlot(RMA, 
                    Logaritmic = "y", 
                    Title="RMA normalization")
BP_fRMA =  BoxPlot(fRMA, 
                    Logaritmic = "y", 
                    Title="fRMA normalization")

### PCA plots
RMA_BG_GeneDF = na.omit(List2DF(RMA_BG_GeneExp))
RMA_GeneDF = na.omit(List2DF(RMA_GeneExp))
fRMA_GeneDF = na.omit(List2DF(fRMA_GeneExp))

PhenoData = PhenoData[which(rownames(PhenoData) %in% colnames(RMA_GeneDF)),]
fRMAPheno = PhenoData[which(rownames(PhenoData) %in% colnames(fRMA_GeneDF)),]

PCA_RMA_BG = PCAPlot(RMA_BG_GeneDF,
                     PhenoData,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") + 
             ggtitle("RMA background correction")
PCA_RMA = PCAPlot(RMA_GeneDF,
                     PhenoData,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") +
          ggtitle("RMA normalization")
PCA_fRMA = PCAPlot(fRMA_GeneDF, 
                     PhenoData,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") + 
           ggtitle("fRMA normalization")
grid_arrange_shared_legend(PCA_RMA_BG, PCA_RMA, PCA_fRMA, ncol=3, nrow = 1)



heatmaply(data.frame(t(RMA_BG_GeneDF) %*% as.matrix(RMA_BG_GeneDF)),
                      col_side_colors = PhenoData[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "RMA background correction")
heatmaply(data.frame(t(RMA_GeneDF) %*% as.matrix(RMA_GeneDF)),
                      col_side_colors = PhenoData[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "RMA normalization")
heatmaply(data.frame(t(fRMA_GeneDF) %*% as.matrix(fRMA_GeneDF)),
                      col_side_colors = fRMAPheno[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "fRMA background correction")

```


## Batch removal

```{r, echo = FALSE, message = FALSE, warning= FALSE}
library(YuGene)

load("B:/OneDrive/Documents/Microarray data/02_Normalized data/01_Colon/BatchCorData_median.RData")

PCA_RMA_QN = PCAPlot(RMA_QN,
                     PhenoData,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") + 
             ggtitle("RMA normalization & quantile normalization")
PCA_RMA_ComBat = PCAPlot(RMA_Combat, 
                     PhenoData,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") +
                 ggtitle("RMA normalization & ComBat batch correction")
PCA_RMA_SVA = PCAPlot(RMA_SVA, 
                     PhenoData,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") +
                 ggtitle("RMA normalization & SVA batch correction")
PCA_RMA_YuGene = autoplot(pca(t(RMA_YuGene)),
                     data = PhenoData, 
                     colour = "Phenotype", 
                     shape = "GEOStudyID") +
                 ggtitle("RMA normalization & YuGene batch correction")
grid_arrange_shared_legend(PCA_RMA_QN,
                           PCA_RMA_ComBat, 
                           PCA_RMA_SVA,
                           PCA_RMA_YuGene,
                           ncol=2,
                           nrow = 2)

PCA_fRMA_QN = PCAPlot(fRMA_QN, 
                     fRMAPheno,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") +
              ggtitle("fRMA normalization & quantile normalization")
PCA_fRMA_ComBat = PCAPlot(fRMA_Combat, 
                     fRMAPheno,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") +
                  ggtitle("fRMA normalization & ComBat batch correction")
PCA_fRMA_SVA = PCAPlot(fRMA_SVA, 
                     fRMAPheno,
                     colorCol = "Phenotype",
                     shapeCol = "GEOStudyID") +
                  ggtitle("fRMA normalization & SVA batch correction")
PCA_fRMA_YuGene = autoplot(pca(t(fRMA_YuGene)),
                     data = fRMAPheno, 
                     colour = "Phenotype", 
                     shape = "GEOStudyID") + 
                  ggtitle("fRMA normalization & YuGene batch correction")
grid_arrange_shared_legend(PCA_fRMA_QN, 
                           PCA_fRMA_ComBat, 
                           PCA_fRMA_SVA,
                           PCA_fRMA_YuGene,
                           ncol=2,
                           nrow = 2)

heatmaply(data.frame(t(RMA_QN) %*% as.matrix(RMA_QN)),
                      col_side_colors = PhenoData[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "RMA normalization & quantile normalization")
heatmaply(data.frame(t(RMA_Combat) %*% as.matrix(RMA_Combat)),
                      col_side_colors = PhenoData[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "RMA normalization & ComBat batch correction")
heatmaply(data.frame(t(RMA_SVA) %*% as.matrix(RMA_SVA)),
                      col_side_colors = PhenoData[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "RMA normalization & SVA batch correction")
heatmaply(data.frame(t(RMA_Combat) %*% as.matrix(RMA_Combat)),
                      col_side_colors = PhenoData[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "RMA normalization & YuGene batch correction")

heatmaply(data.frame(t(fRMA_QN) %*% as.matrix(fRMA_QN)),
                      col_side_colors = fRMAPheno[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "fRMA normalization & quantile normalization")
heatmaply(data.frame(t(fRMA_Combat) %*% as.matrix(fRMA_Combat)),
                      col_side_colors = fRMAPheno[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "fRMA normalization & ComBat batch correction")
heatmaply(data.frame(t(fRMA_SVA) %*% as.matrix(fRMA_SVA)),
                      col_side_colors = fRMAPheno[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "fRMA normalization & SVA batch correction")
heatmaply(data.frame(t(fRMA_YuGene) %*% as.matrix(fRMA_YuGene)),
                      col_side_colors = fRMAPheno[,c("GEOStudyID", "Phenotype")],
                      seriate = "mean", 
                      row_dend_left = TRUE,
                      plot_method = "plotly",
                      main = "fRMA normalization & YuGene batch correction")
```

## Add gene name and uniprot ID

```{r}
#source("Annotate.R")
#fRMA_Combat = Symbol2UniprotID(fRMA_Combat)

```


# Differential gene analysis
```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(limma)

design = model.matrix(~0 + factor(PhenoData$Phenotype))
colnames(design) = unique(as.character(PhenoData$Phenotype))

fit = lmFit(fRMA_Combat, design)
contrMatrix = makeContrasts(Normal-Intermediate, Intermediate-Cancer, Normal-Cancer, levels=design)
fit2 = contrasts.fit(fit, contrMatrix)
fit2 = eBayes(fit2)

tested = topTable(fit2, adjust = "fdr", sort.by = "B", number = Inf, coef = 1)
NormIntermDE = tested[tested$adj.P.Val <0.01,]

tested = topTable(fit2, adjust = "fdr", sort.by = "B", number = Inf, coef = 2)
IntermCancerDE = tested[tested$adj.P.Val <0.01,]

tested = topTable(fit2, adjust = "fdr", sort.by = "B", number = Inf, coef = 3)
NormCancerDE = tested[tested$adj.P.Val <0.01,]

head(NormIntermDE)
head(IntermCancerDE)
head(NormCancerDE)


```

# Functional annotation of DE genes

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(clusterProfiler)
library(org.Hs.eg.db)

NormalIntermGO = enrichGO(gene = rownames(NormIntermDE),
                          universe = rownames(fRMA),
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.01,
                          qvalueCutoff = 0.05)
NormalIntermGO = dropGO(NormalIntermGO, level = 1)
IntermCancerGO = enrichGO(gene = rownames(IntermCancerDE),
                          universe = rownames(fRMA),
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.01,
                          qvalueCutoff = 0.05)
IntermCancerGO = dropGO(IntermCancerGO, level = 1)
NormalCancerGO = enrichGO(gene = rownames(NormCancerDE),
                          universe = rownames(fRMA),
                          OrgDb = org.Hs.eg.db,
                          keyType = "SYMBOL",
                          ont = "BP",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.01,
                          qvalueCutoff = 0.05)
NormalCancerGO = dropGO(NormalCancerGO, level = 1)

heatplot(NormalIntermGO)
heatplot(IntermCancerGO)
heatplot(NormalCancerGO)

```

